State-space analysis of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms

### 2023-11-30 manuscript/manuscript_2/paper.tex
version passed to AM

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is Kalman-tracked concurrently with the modulation induced by a single-source gravitational wave in the pulse times of arrival. The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively. It is shown how to extend previous state-space analyses to include the pulsar term. Astrophysically representative software injections in Gaussian measurement noise evidence that for a loud GW source with characteristic wave strain $h_0 = 10^{-12}$ the inclusion of the pulsar terms corrects for biases in the parameter estimates and decreases the root-mean-square relative error in the estimated one-dimensional posteriors by up to 99\%. For a quiet GW source with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in the parameter accuracy since the system is heavily noise dominated. However, including the pulsar terms decreases the minimal detectable strain by \textcolor{red}{Y percent} .
\end{abstract}

@@

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
Nanohertz gravitational waves (GWs) are produced by the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022}. A nHz GW influences the time of arrival (TOA) at the Earth of radio pulses from pulsars. A pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple pulsars. In this way it is possible to detect GWs at frequencies that are prohibitively low for terrestrial interferometers. Consilient evidence from PTA observations has been presented \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X} for a stochastic GW background, which arises from the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. SMBHBs which are sufficiently close and massive are perhaps individually resolvable, though no definitive detection is yet claimed \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}. \newline

@@

\cite{KimpsonPTA} (K24 hereafter) introduced a novel, state-space method for the detection and parameter estimation of continuous GWs from individual SMBHBs, that self-consistently tracks the intrinsic timing noise in PTA pulsars \citep[e.g.][]{Shannon2010,Lasky2015,Caballero2016,Goncharov2021} and disentangles it from GW-induced TOA modulations. The method described in K24 follows the example of standard  PTA analyses and also drops the pulsar term. In this paper we undertake two key tasks. Firstly we extend the methods presented in K24 to include the pulsar term. We show how to reparametrise the state-space model in terms of an additional pulsar-specific phase term. We demonstrate how the static GW parameters and the additional pulsar phase term can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}. Secondly we quantify how the inclusion of the pulsar term affects (i) the accuracy with which the static GW parameters can be estimated, and (ii) the minimal detectable GW strain, compared to the case where the pulsar terms are omitted.  \newline

@@

\subsection{Spin evolution} \label{sec:psr_frequency}
The intrinsic pulsar spin frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ , 
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm em}^{(n)}$ is the deterministic evolution, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies
\begin{align}
	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'}(t - t') \ .	\label{eq:xieqn}
\end{align}
In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ which parametrizes the noise amplitude and gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}. PTAs are typically composed of millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$. Consequently, 	$f_{\rm em}^{(n)}(t)$ can be approximated accurately by 
\begin{equation}
	f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} 
where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. \newline

@@

\subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
In the presence of a GW, the intrinsic pulsar spin frequency for the $n$-th pulsar is related to the radio pulse frequency measured by an observer on Earth via a measurement equation:
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}(t)\ ,
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ ,	\label{eq:vareps}
\end{align}
where $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
	\label{eq:g_func_trig}
\end{align}
where $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$, $\Omega$ is the constant angular frequency of the GW, $\boldsymbol{n}$ is a unit vector specifying the direction of propagation of the GW, $H_{ij}$ is the spatial part of the gravitational plane wave amplitude tensor, and $\Phi_0$ the phase offset of the GW with respect to some reference time. \newline

@@

Regarding point (ii) whilst in practice the pulsar locations vary with respect to the Earth, pulsar TOAs are defined relative to the Solar System barycentre (SSB).The barycentering correction is typically applied when generating TOAs, e.g. with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series. Some pulsars, including some PTA pulsars, do have non-negligible proper motions of order $10^2$ km s$^{-1}$ after the barycentering corrections have been applied \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this paper.

@@

\section{Inclusion of the pulsar term}\label{sec:pulsar_term}
K24 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection. That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}(t)$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}, $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$, and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
For our purposes ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$. Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is solely a function of $\boldsymbol{\theta}$, i.e. $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}) = \mathcal{L}(\boldsymbol{\theta})$. Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$. In the context of PTA searches for GWs, we refer the reader to Appendix  \ref{sec:kalman} for a review of Kalman filtering, Appendix \ref{sec:nested_sampling} for a review of nested sampling, Appendix \ref{sec:workflow} for an overview of the combined Kalman filter - nested sampling workflow. \newline

@@

Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases which cannot be coherently summed and are instead generally considered as a source of self-noise. The pulsar term depends on $d^{(n)}$ which is generally poorly constrained by electromagnetic observations, with uncertainties much greater than the target GW wavelength. It is therefore not possible to set the phase of the pulsar term. Consequently the pulsar term is often dropped in standard PTA analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}. Dropping the pulsar term is also convenient computationally because it reduces the dimensionality of the parameter space; the $N$ values of $d^{(n)}$ are not inferred, making the parameter estimation computationally cheaper. Within the state-space model described in Section \ref{sec:2}, dropping the pulsar terms equates to using a modified measurement equation, c.f. Equations \eqref{eq:measurement} and \eqref{eq:g_func_trig},
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}_{\rm Earth}(t) + \varepsilon^{(n)}(t) \ , 
	\label{eq:measuremen_earth}
\end{equation}
with
\begin{equation}
	g^{(n)}_{\rm Earth}(t) = 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \cos(-\Omega t +\Phi_0)  \ .
	\label{eq:g_func_trig_earth}
\end{equation}
However, dropping the pulsar term leads to biases in the parameter estimation, in particular the source localisation, and a reduced detection probability \citep{Zhupulsarterms,Chen2022,KimpsonPTA}. It is therefore important to remove this limitation and include the pulsar term in PTA data analysis pipelines.

@@

\section{Validation with synthetic data}\label{sec:rep_example1}
In this section we apply the analysis scheme of K24 while including the pulsar terms using the parametrisation described in Section \ref{sec:pulsar_term2}. As outlined in Section \ref{sec:pulsar_term}, the analysis scheme consists of a Kalman filter to track the state evolution of the intrinsic pulsar spin frequency $f_{\rm p}^{(n)}(t)$ which provides a likelihood for nested sampling. Nested sampling then returns estimates of $\boldsymbol{\theta'}$ and $\mathcal{Z}$. We refer the reader to Appendix \ref{sec:kalman} and K24 for details of the Kalman filter - nested sampling scheme. We apply the analysis scheme to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source. The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper. The goal of this section is to validate the analysis scheme when the pulsar terms are included in the inference model. \newline

In Section \ref{sec:synthetic_data} we demonstrate how to generate synthetic data, namely noisy frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$, with which to validate the scheme. In Section \ref{sec:pe_and_ms} we apply the analysis scheme to the synthetic data to estimate $\boldsymbol{\theta'}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$. The scheme is validated on multiple noise realisations of the pulsar process noise and the measurement noise in order to test the scheme more strongly and quantify the  irreducible ``cosmic'' variance in the parameter estimates. In Section \ref{sec:parameter_space} we extend the tests of Section  \ref{sec:pe_and_ms}
to cover a broader parameter space and consider a range of astrophysically relevant SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$.

@@

In this paper we use the same values for $\boldsymbol{\theta}_{\rm psr}$ as in K24, i.e. we adopt the $N=47$ millisecond pulsars (MSPs) in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}, i.e. $N = 47$. We assume all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$ over a 10 year period. For each pulsar fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}^{(n)}_{\rm em}(t_1)$ are obtained via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. We identify $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and its time derivative respectively, as quoted in catalogues. No direct measurements exist for $\gamma^{(n)}$, though the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}. In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. Similarly, no direct measurements exist for $\sigma^{(n)}$.   We relate the process noise in the frequency $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week such that
\begin{eqnarray}
	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
\end{eqnarray}
where, following K24, we use the empirical timing noise model for MSPs from \cite{Shannon2010ApJ...725.1607S} in order to set physically reasonable values on $\sigma_{\rm TOA}^{(n)}$.  The median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134. \newline

@@

%images/corner_lowSNR_chiF
	\begin{figure}
	\includegraphics[width=\columnwidth, height =\columnwidth ]{images/corner_lowSNR_chi_n4000}
	\caption{Same as Figure \ref{fig:corner_plot_1}	but for five static parameters of $\chi^{(n)}$. Consistent unimodal posteriors are obtained across the nine noise realisations for four out of the five displayed parameters. No consistent posteriors are obtained for $\chi^{(2)}$ (third row, third column), because this pulsar has a similar sky position to that of the synthetic GW source. The known, injected value lies within the 90\% credible interval for 38 out of the 50 combinations of five parameters and nine noise realizations (i.e. $76 \%$). }
	\label{fig:corner_plot_3}
\end{figure}

>>>>

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked using a Kalman filter concurrently with the modulation induced by a single-source gravitational wave in the pulse times of arrival. The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively. It is shown how to extend previous state-space analyses to include the pulsar term. Astrophysically representative software injections in Gaussian measurement noise evidence that for a loud GW source with characteristic wave strain $h_0 = 10^{-12}$ the inclusion of the pulsar terms corrects for biases in the parameter estimates and decreases the root-mean-square relative error in the estimated one-dimensional posteriors by up to 99\%. For a quiet GW source with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in the parameter accuracy since the system is heavily noise dominated. However, including the pulsar terms decreases the minimal detectable strain by $8\%$ .
\end{abstract}

@@

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
Nanohertz gravitational waves (GWs) are produced by the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022}. A nHz GW can influence the time of arrival (TOA) at the Earth of radio pulses from pulsars. A pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple pulsars. In this way it is possible to detect GWs at frequencies that are prohibitively low for terrestrial interferometers. Consilient evidence from PTA observations has been presented \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X} for a stochastic GW background, which arises from the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. SMBHBs which are sufficiently close and massive are perhaps individually resolvable, though no definitive detection is yet claimed \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}. \newline

@@

\cite{KimpsonPTA} (K24 hereafter) introduced a novel, state-space method for the detection and parameter estimation of continuous GWs from individual SMBHBs, that self-consistently tracks the intrinsic timing noise in PTA pulsars \citep[e.g.][]{Shannon2010,Lasky2015,Caballero2016,Goncharov2021} and disentangles it from GW-induced TOA modulations. The method described in K24 follows the example of standard  PTA analyses and also drops the pulsar term. In this paper we undertake two key tasks. Firstly we extend the methods presented in K24 to include the pulsar term. We show how to reparametrise the state-space model in terms of an additional pulsar-specific phase term. We demonstrate how the static GW parameters and the additional pulsar phase term can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}. Secondly, we quantify how the inclusion of the pulsar term affects (i) the accuracy with which the static GW parameters can be estimated, and (ii) the minimal detectable GW strain, compared to the case where the pulsar terms are omitted.  \newline

@@

\subsection{Spin evolution} \label{sec:psr_frequency}
The intrinsic pulsar spin frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ , 
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm em}^{(n)}$ is the deterministic evolution, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative where an overdot denotes a derivative with respect to $t$, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies
\begin{align}
	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'}(t - t') \ .	\label{eq:xieqn}
\end{align}
In Equation \eqref{eq:xieqn}, $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ which parametrizes the noise amplitude and gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}. PTAs are typically composed of millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$. Consequently, 	$f_{\rm em}^{(n)}(t)$ can be approximated accurately by 
\begin{equation}
	f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} 
where $t_1$ labels the time of the first TOA. \newline

@@

\subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
In the presence of a GW, the intrinsic pulsar spin frequency for the $n$-th pulsar is related to the radio pulse frequency measured by an observer on Earth via a measurement equation
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}(t)\ ,
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ .	\label{eq:vareps}
\end{align}
In Equation \eqref{eq:vareps}, $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars. In Equation \eqref{eq:measurement} the measurement function $g^{(n)}(t)$ is
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ ,
	\label{eq:g_func_trig}
\end{align}
where $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$, $\Omega$ is the constant angular frequency of the GW, $\boldsymbol{n}$ is a unit vector specifying the direction of propagation of the GW, $H_{ij}$ is the spatial part of the gravitational plane wave amplitude tensor, and $\Phi_0$ the phase offset of the GW with respect to some reference time. \newline

@@

Regarding point (ii) whilst in practice the pulsar locations vary with respect to the Earth, pulsar TOAs are defined relative to the Solar System barycentre (SSB). The barycentering correction is typically applied when generating TOAs, e.g. with {\sc tempo2} \citep{tempo2} and related timing software, and is inherited by the frequency time series. Some pulsars, including some PTA pulsars, do have non-negligible proper motions of order $10^2$ km s$^{-1}$ after the barycentering corrections have been applied \citep[e.g.][]{10.1093/mnras/sty3390}, but we do not consider this effect in this paper.

@@

\section{Inclusion of the pulsar term}\label{sec:pulsar_term}
K24 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection. That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}(t)$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}, $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$, and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
For our purposes ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$. Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is solely a function of $\boldsymbol{\theta}$, i.e. $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}) = \mathcal{L}(\boldsymbol{\theta})$. Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$. Within the context of PTA searches for GWs, we refer the reader to Appendix  \ref{sec:kalman} for a review of Kalman filtering, Appendix \ref{sec:nested_sampling} for a review of nested sampling and Appendix \ref{sec:workflow} for an overview of the combined Kalman filter - nested sampling workflow. \newline

@@

Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases which cannot be coherently summed and are instead generally considered as a source of self-noise. The pulsar term depends on $d^{(n)}$ which is generally poorly constrained by electromagnetic observations, with uncertainties much greater than the target GW wavelength. Consequently the pulsar term is often dropped in standard PTA analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}. Dropping the pulsar term is also convenient computationally because it reduces the dimensionality of the parameter space; the $N$ values of $d^{(n)}$ are not inferred, making the parameter estimation computationally cheaper. Within the state-space model described in Section \ref{sec:2}, dropping the pulsar terms equates to using a modified measurement equation, c.f. Equations \eqref{eq:measurement} and \eqref{eq:g_func_trig},
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}_{\rm Earth}(t) + \varepsilon^{(n)}(t) \ , 
	\label{eq:measuremen_earth}
\end{equation}
with
\begin{equation}
	g^{(n)}_{\rm Earth}(t) = 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \cos(-\Omega t +\Phi_0)  \ .
	\label{eq:g_func_trig_earth}
\end{equation}
However, dropping the pulsar term leads to biases in the parameter estimation, in particular the source localisation, and a reduced detection probability \citep{Zhupulsarterms,Chen2022,KimpsonPTA}. It is therefore important to remove this limitation and include the pulsar term in PTA data analysis pipelines.

@@

\section{Validation with synthetic data}\label{sec:rep_example1}
In this section we apply the analysis scheme of K24 while including the pulsar terms using the parametrisation described in Section \ref{sec:pulsar_term2}. As outlined in Section \ref{sec:pulsar_term}, the analysis scheme consists of a Kalman filter to track the state evolution of the intrinsic pulsar spin frequency $f_{\rm p}^{(n)}(t)$ which provides a likelihood for nested sampling. Nested sampling then returns estimates of $\boldsymbol{\theta'}$ and $\mathcal{Z}$. We refer the reader to Appendices \ref{sec:kalman} -- \ref{sec:workflow} and K24 for details of the Kalman filter - nested sampling scheme. We apply the analysis scheme to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source. The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper. The goal of this section is to validate the state-space analysis scheme when the pulsar terms are included in the inference model. \newline

In Section \ref{sec:synthetic_data} we demonstrate how to generate synthetic data, namely noisy frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$, with which to validate the scheme. In Section \ref{sec:pe_and_ms} we apply the analysis scheme to the synthetic data to estimate $\boldsymbol{\theta'}$ for a  particular, arbitrary, representative choice of $\boldsymbol{\theta}_{\rm gw}$, i.e. the SMBHB system. The scheme is validated on multiple noise realisations of the pulsar process noise $\xi^{(n)}(t)$ and the detector measurement noise $\varepsilon^{(n)}(t)$ in order to test the scheme more strongly and quantify the  irreducible ``cosmic'' variance in the parameter estimates. In Section \ref{sec:parameter_space} we extend the tests of Section  \ref{sec:pe_and_ms}
to cover a broader parameter space and consider a range of astrophysically relevant SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$.

@@

In this paper we use the same values for $\boldsymbol{\theta}_{\rm psr}$ as in K24, i.e. we adopt the $N=47$ millisecond pulsars (MSPs) in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}. We assume all pulsars are observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$ over a 10 year period. For each pulsar fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}^{(n)}_{\rm em}(t_1)$ are obtained via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. We identify $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and its time derivative respectively, as quoted in catalogues. No direct measurements exist for $\gamma^{(n)}$, though the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}. In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. Similarly, no direct measurements exist for $\sigma^{(n)}$.   We relate the process noise in the frequency $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week such that
\begin{eqnarray}
	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
\end{eqnarray}
where, following K24, we use the empirical timing noise model for MSPs from \cite{Shannon2010ApJ...725.1607S} in order to set physically reasonable values on $\sigma_{\rm TOA}^{(n)}$.  The median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134. \newline

@@

%images/corner_lowSNR_chiF
	\begin{figure}
	\includegraphics[width=\columnwidth, height =\columnwidth ]{images/corner_lowSNR_chi_n4000}
	\caption{Same as Figure \ref{fig:corner_plot_1}	but for five static parameters of $\chi^{(n)}$. Consistent unimodal posteriors are obtained across the ten noise realisations for four out of the five displayed parameters. No consistent posteriors are obtained for $\chi^{(2)}$ (third row, third column), because this pulsar has a similar sky position to that of the synthetic GW source. The known, injected value lies within the 90\% credible interval for 38 out of the 50 combinations of five parameters and ten noise realizations (i.e. $76 \%$). }
	\label{fig:corner_plot_3}
\end{figure}
